---
title: "`clusterplot`: A package for graphical exploration of big datasets"
author: "Rebecca Barter"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
header-includes: \usepackage{graphicx}
vignette: >
  %\VignetteIndexEntry{Using the modelfit package for assessing the fit of a model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Introduction 

The `clusterplot` package was developed as a data exploration plot designed to visually explore big datasets. The highly customisable plot is generated by combining a clustered heatmap with scatterplots with the aim of both visualising clusters in the data as well as how the relationship between our response variable and other variables of interest differs for different clusters.



--------


## Downloading and installing the package


```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE
)
```



<!--# Using the modelfit package for assessing the fit of a model. -->

The goal of this guide is to help you understand how and why to use the `clusterplot` package in R to assess your data model of interest. First, you need to download and install the package. This can be done using the `devtools` package. If you have not yet done so, you will need to download it. The `devtools` package allows the user to download and install packages hosted on github pages, such as this one. To do this (for most non-Windows systems), you need to type the following code into your R console:


```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("rlbarter/clusterplot")
```




Next, you can load the `clusterplot` library into your workspace:

```{r}
library(clusterplot)

```


--------

## The contents of the plot

As mentioned above, the primary aim of this package is to produce a plot that can be used to diagnose areas of your data where your model may be performing sub-optimally (and to gives clues about possible improvements), as well as provide a means for visualizing the structure of your data. The plot consists of two elements:

1. *a clustered heatmap*: a clustered version of the $(n \times p)$ design matrix ${\bf X}$).

1. *a scatterplot*: a scatterplot to the right of the heat map. 


--------------

## Things you should keep in mind
The heatmap in the center of the plot is a $(p \times n)$ design matrix, ${\bf X}$, in which the observations have been clustered into a number of distinct clusters (displayed on the left and bottom labels). In particular, the $(i,j)$th entry in the heatmap corresponds to the $i$th observation for the $j$th variable in $X$. The colour scale is presented below the heatmap. 


--------

## Usage

The package consists of one main function: `clusterplot`. 

The `clusterplot` function takes data objects, the most important of which are `X` (the heatmap matrix) and `yr` (a vector of values to be plotted to the right of the heatmap). For example, the plot could be generated for the `iris` dataset as follows
```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
# define a linear model and isolate coefficients for each variable
iris.coef <- lm(Petal.Length ~ Sepal.Length + Sepal.Width + Petal.Width, data = iris)$coef

# generate the plot:
# a heatmap of the matrix X (the design matrix for our linear model)
# plot a scatterplot of yr on the right (the response for our linear model)
# cluster the rows by Species
# plot a barplot of yt on the top (the coefficients - take abs value and remove intercept)
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            membership.rows = iris[,"Species"],
            yt = abs(iris.coef[-1]),
            yt.plot.type = "bar",
            yt.axis.name = "absolute coefficient")
```

This plot shows us a heatmap of the design matrix, $X$, which consists of 150 rows (the first 10 of which are shown below) and 3 variables (the first three columns of the table below). We also have a species variable (the fourth column of the table below) that we are using as our cluster membership identifier. Finally, on the right hand side of the plot we are plotting a fourth variable `Sepal.Length`. We can see from this plot that the virginica species has longer petals and sepals than setose which has short sepals and particularly short petal width.
```{r, echo = FALSE, results = "asis"}
pander::pandoc.table(iris[1:10,c("Sepal.Width","Sepal.Length","Petal.Width", "Species")])
```



Further, if you are one of those people who prefer their design matrix to be $n \times p$ instead of $p \times n$, we can still accomodate you as the plot below shows! For example, we are plotting a heatmap of the transpose of $X$, as well as the `Sepal.Length` variable above (`yt` corresponds to the variable to be plotted on *top* of the heatmap). Note that below we must specify `cluster.rows = FALSE`, since the default is to cluster the rows but not the columns (we will discuss the default settings below).\

```{r, fig.align='center',fig.height = 7, fig.width = 5.5,  fig.show = 'hold'}

clusterplot(X = t(iris[,c("Sepal.Width","Sepal.Length","Petal.Width")]), 
            yt = iris[,"Petal.Length"], 
            yt.axis.name = "Petal.Length",
            membership.cols = iris[,"Species"], 
            cluster.rows = FALSE)
```





--------

## Customizing the plot
Below we continue with the iris example and show how to use the numerous options for customizing the plot


### Clustering options (row *and* column)

In the `iris` example above, we have specified our own cluster vector (the `Species` variable which consists of three classes: Setose, Versicolor and Virginica). However, if you do not have a cluster membership vector in mind (or cannot be bothered to generate one using your clustering algorithm of preference), never fear! The `clusterplot` function has in-built clustering options for performing K-means (the default if a membership vector is not supplied) and hierarchical clustering (specify `clustering.method = "hierarchical"`) on the rows and/or columns.\\



```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
set.seed(134)
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            n.clusters.rows = 3)

```


Again we note that, by default, if no membership vectors are supplied, the default is to cluster the rows but not the columns. As a result, the number of row clusters must be supplied (below we have `n.clusters.rows = 3`), otherwise an error will be produced. If you wish to cluster the columns also, you could specify a number of column clusters using the `n.cluster.col` argument.



```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
set.seed(134)
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            n.clusters.rows = 3,
            n.clusters.col = 2)
```

If you wanted to keep the variable names as the labels, you could simply specify `bottom.heat.label = "variable"`.

```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
set.seed(134)
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            n.clusters.rows = 3,
            n.clusters.col = 2,
            bottom.heat.label = "variable")
```


Moreover, if you have used the in-built clustering algorithms, you can obtain the clustering membership vector by saving the `clusterplot` object and acessing the `membership` element (below we hide the plot itself, which is the same as the previous plot, by specifying `print.plot = F`).

```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
set.seed(134)
plot <- clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            n.clusters.rows = 3,
            print.plot = F)
plot$membership.rows
```


We note that the k-means clustering corresponds mostly to the Species variable.


### Layout



It is not necessary to have all components at all times in your plot. For example, if you do not supply a `yt` vector, no top scatterplot is provided. Similarly if you do not supply a  `yr` vector. Further options include:

* You can remove the legend:


```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            membership.rows = iris[,"Species"],
            legend = FALSE)
```


* You can remove the left cluster/variable labels

```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            membership.rows = iris[,"Species"],
            left.heat.label = "none")
```

* You can remove the bottom cluster/variable labels

```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"], 
            yr.axis.name = "Petal.Length",
            membership.rows = iris[,"Species"],
            bottom.heat.label = "none")
```


* You can remove the scatterplot axis (similarly for the top axis)

```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis = F)
```

* You can add a title

```{r, fig.align='center',fig.height = 5, fig.width = 6,  fig.show = 'hold'}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            title = "I'm a title!")
```




* You can add white padding (in cm) around the plot (the default is 2cm):

```{r, fig.align = 'center', fig.height = 5, fig.width = 6}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            padding = 2.5)
```





### Colour palattes
The default colour palates are as shown in the above examples, however the user can specify their own colour palates:

 * There are a number of optional colour schemes for the heatmap including "red" (the default), "purple", "blue" (shown below), "grey" and "green". These can be specified using the `heat.col.scheme` argument.
 
 
```{r, fig.align = 'center', fig.height = 5, fig.width = 6}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            heat.col.scheme = "blue")
```

* You can also specify your own colour scheme for the heatmap using the `heat.pal` argument. You can specify as many colours as you want. You can also use the `heat.pal.values` argument to specify the position of each colour in the scale from 0 to 1.

 
```{r, fig.align = 'center', fig.height = 5, fig.width = 6}
clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            heat.pal = c("yellow", "brown", "blue"))
```


*  You can specify the colour of the points in the scatterplot (replace `yr` with `yt` for the top scatterplot). Note that the order of the points in the `yr.col` vector correspond to the order of the points in `yr`.


```{r, fig.align = 'center', fig.height = 5, fig.width = 6}
col.vec <- rep("black", 150)
col.vec[50:70] <- "red"

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            yr.col = col.vec)
```

* You can specify the cluster label (and variable label) colors. (The equivalent for the bottom labels can be specified using `bottom.label.pal` and `bottom.label.text.col`)



```{r, fig.align = 'center', fig.height = 5, fig.width = 6}

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            left.label.pal = c("blue","red","purple"),
            left.label.text.col = c("white","yellow","white"))
```





### Sizing

There are a number of options to change the text size, point size and legend size in the plot. For example, you can specify the following arguments in the `clusterplot` function:


* You can change the legend size


```{r, fig.align = 'center', fig.height = 5, fig.width = 6}

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            legend.size = 4)
```


* You can change the cluster/variable label size (the default is 0.1)

```{r, fig.align = 'center', fig.height = 5, fig.width = 6}

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            bottom.label.size = 0.5,
            bottom.text.angle = 90,
            left.label.size = 0.4,
            left.text.angle = 0)
```


* You can change the cluster/variable label text size (the default is 5)


```{r, fig.align = 'center', fig.height = 5, fig.width = 6}

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            left.text.size = 7,
            bottom.text.size = 2)
```


* You can change the scatterplot axis size and the axis label size (the default for both is 10). 


```{r, fig.align = 'center', fig.height = 5, fig.width = 6}

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            yr.axis.size = 15,
            yr.axis.name.size = 20)
```

* you can change the point size

```{r, fig.align = 'center', fig.height = 5, fig.width = 6}

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            yr.point.size = 3.5)
```


### Ordering

* you can change order of observations within each cluster. For example, we can order the observations by Sepal.Width

```{r, fig.align = 'center', fig.height = 5, fig.width = 6}

clusterplot(X = iris[,c("Sepal.Width","Sepal.Length","Petal.Width")], 
            yr = iris[,"Petal.Length"],
            membership.rows = iris[,"Species"],
            yr.axis.name = "Petal.Length",
            order.rows = order(iris$Sepal.Width))
```


--------

## The end

Thanks for using the package! I hope you find it helpful in your data exploration adventures. The development page can be found at <https://github.com/rlbarter/clusterplot>. For pull requests and suggestions please follow the standard protocol. For pressing questions or comments feel free to email me at rebeccabarter@berkeley.edu.



--------
